---
title: "Assignment 2: Coding Basics"
author: Claire Mullaney
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A02_CodingBasics.Rmd") prior to submission.

The completed exercise is due on Tuesday, January 21 at 1:00 pm.


## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
seq100 <- seq(1, 100, 4) #Creating a sequence of numbers that goes from 1 to 100, counting by 4

#2. 
mean(seq100) #Calculating the mean of this sequence
median(seq100) # Calculating the median of this sequence

#3. 

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5
##(a)
students <- c("Claire", "Nikki", "Masha", "Maggie") #character vector containing the names of students

##(b)
test.scores <- c(97, 85, 43, 64) #numeric vector containing test scores

##(c)
pass.fail <- c(T, T, F, T) #logical vector indicating if students passed the test (T) or failed the test (F)

#7
student.scores <- data.frame(students, test.scores, pass.fail) #Combining vectors into a data frame

#8
names(student.scores) <- c("Students", "Test Scores", "Score >= 50") #renaming the columns of the data frame
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix is a 2D structure contains elements that are all of the same type, while a data frame can contain data of different types (numeric, character, logical) 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10.
#function to find passing test scores using ifelse
pass.ifelse <- function(x) {
  ifelse( x >= 50, T, F)
}
#function to find passing test scores using if and lse
pass.if.else <- function(x) {
  if(x >= 50)
    print(TRUE)
  else
    print(FALSE)
}

#11
#applying functions to the test score vector above
pass.ifelse(test.scores)
pass.if.else(test.scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: `ifelse` worked, while `if` and `else` did not. `ifelse` preserved the shape of the original vector, evaluating each component to return the same number of T/F values as there were numbers. Using `if` and `else` only allowed one value from the vector to be evaluated; it is not set up to loop through the entire vector.
