---
title: "Assignment 5: Data Visualization"
author: "Claire Mullaney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 11 at 1:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse and cowplot packages. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (tidy and gathered) and the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
#Verifying directory, loading packages, and uploading files
getwd()

library(ggplot2)
library(viridis)
library(RColorBrewer)
library(tidyverse)
library(cowplot)


Peter_Paul_Nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
Peter_Paul_Nutrients_Gathered <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
Niwot_Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
#Looking at column names, checking the class of each date column, and converting those date columns to dates

###Peter_Paul_Nutrients
colnames(Peter_Paul_Nutrients)
class(Peter_Paul_Nutrients$sampledate)

Peter_Paul_Nutrients$sampledate <- 
  as.Date(Peter_Paul_Nutrients$sampledate, 
          format = "%Y-%m-%d")

###Peter_Paul_Nutrients_Gathered
colnames(Peter_Paul_Nutrients_Gathered)
class(Peter_Paul_Nutrients_Gathered$sampledate)

Peter_Paul_Nutrients_Gathered$sampledate <- 
  as.Date(Peter_Paul_Nutrients_Gathered$sampledate, 
          format = "%Y-%m-%d")

###Niwot_Litter
colnames(Niwot_Litter)
class(Niwot_Litter$collectDate)

Niwot_Litter$collectDate <- 
  as.Date(Niwot_Litter$collectDate, 
          format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
mytheme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right") +

theme_set(mytheme)

#Can create a custom theme or use one that already exists
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus by phosphate, with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}

tp_vs_po4 <- ggplot(Peter_Paul_Nutrients, aes(y = tp_ug, x = po4, color = lakename)) +
  geom_point() +
  geom_smooth(method=lm, se = FALSE, color = "black", size = 0.5) +
  xlim(0, 50) +
  ylim(0, 150) +
  labs(y = expression(paste("TP ( ", mu, "g/L)")), x = expression(paste("PO"[4]* " (", mu, "g/L)")), color = "Lake Name") +
  scale_color_viridis(discrete = TRUE)

print(tp_vs_po4)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}

#Converting month variables to factors (rather than numeric variables) so they can be used to break out the data in each boxplot
Peter_Paul_Nutrients$month <- as.factor(Peter_Paul_Nutrients$month)

Peter_Paul_Nutrients_Gathered$month <- as.factor(Peter_Paul_Nutrients_Gathered$month)

#Constructing boxplots

temp_box <- ggplot(Peter_Paul_Nutrients, 
                   aes(y = temperature_C, x = month, 
                       color = lakename)) +
  geom_boxplot() + 
  labs(y = "Temperature (C)", x = "Month", color = "Lake Name") +
  scale_color_manual(values = c("#d7191c", "#2c7bb6"))

print(temp_box)
##########
tp_box <- ggplot(Peter_Paul_Nutrients, 
                 aes(y = tp_ug, x = month, color = lakename)) +
         geom_boxplot() + 
  labs(y = expression(paste("TP ( ", mu, "g/L)")), 
       x = "Month", color = "Lake Name") +
  scale_color_manual(values = c("#d7191c", "#2c7bb6"))
  
print(tp_box)
##########
tn_box <- ggplot(Peter_Paul_Nutrients, 
                 aes(y = tn_ug, x = month, color = lakename)) +
         geom_boxplot() +
    labs(y = expression(paste("TN ( ", mu, "g/L)")), x = "Month",
         color = "Lake Name") +
  scale_color_manual(values = c("#d7191c", "#2c7bb6"))

print(tn_box)

#Create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

temp_tp_tn <- plot_grid(temp_box + theme(legend.position="none"), tp_box + theme(legend.position="none"), tn_box + theme(legend.position="none"), nrow = 3, align = "hv", rel_heights = c(1.25, 1))

legend <- get_legend(temp_box + theme(legend.box.margin = margin(0, 0, 0, 12)))

plot_grid(temp_tp_tn, legend, rel_widths = c(3, .4))

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Median temperature and temperature ranges increase from May through August and decrease from September through November. Peter Lake consistently has lower median temperatures, and often wider temperature ranges, than Paul Lake (except in the month of October). The total amount of Nitrogen in the lake follows a similar pattern, with the median amount increasing for both lakes from May through August and starting to decrease in September. Peter Lake often has larger median amounts of nitrogen, a larger interquartile range, and more positively skewed than Paul Lake. The total amount of phosphorous in the lakes appears to increase from May through September. Peter Lake has higher median amounts of total phosphorus, and the distributions for Paul Lake are more positively skewed. There are many upper outliers in the distributions for both total nitrogen and total phosphorus in each month.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
drymass_vs_date <- ggplot(subset(Niwot_Litter, 
                                 functionalGroup == "Needles")) +
  aes(y = dryMass, x = collectDate, color = nlcdClass) +
  geom_point() +
  labs(y = "Dry Mass (g)", x = "Collection Date", color = "NLCD Class") +
  scale_color_hue(labels = c("Evergreen Forest", "Grassland Herbaceous", "Shrub Scrub")) +
  scale_color_viridis(discrete = TRUE)

print(drymass_vs_date)

#7
nlcd_labels <- c(evergreenForest = "Evergreen Forest", grasslandHerbaceous = "Grassland Herbaceous", shrubScrub ="Shrub Scrub")

drymass_vs_date_fac <- ggplot(subset(Niwot_Litter, 
                                     functionalGroup == "Needles")) +
  aes(y = dryMass, x = collectDate) +
  geom_point() +
  facet_wrap(vars(nlcdClass), labeller = as_labeller(nlcd_labels)) +
  labs(y = "Dry Mass (g)", x = "Collection Date")


print(drymass_vs_date_fac)
```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think plot 6 is more effective. I find it easier to compare all three NLCD classes within each year when the points are all on the same graph and separated by color. When each NLCD class is a different section of the faceted graph, scanning the graph to compare these different classes within years is challenging. Even though there is no need to separate the classes by color because they are on separate graphs, the lack of a clear visual distinction among the points makes comparing the different classes difficult.