---
title: "Assignment 5: Data Visualization"
author: "Claire Mullaney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 11 at 1:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse and cowplot packages. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (tidy and gathered) and the processed data file for the Niwot Ridge litter dataset. 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
#Verifying directory, loading packages, and uploading files
getwd()

library(ggplot2)
library(tidyverse)
library(cowplot)


Peter_Paul_Nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")
Peter_Paul_Nutrients_Gathered <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
Peter_Paul_ChemistryPhysics <-
  read.csv("./Data/Processed/NTL-LTER_Lake_ChemistryPhysics_PeterPaul_Processed.csv")
Niwot_Litter <- read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2
#Looking at column names, checking the class of each date column, and converting those date columns to dates

###Peter_Paul_Nutrients
colnames(Peter_Paul_Nutrients)
class(Peter_Paul_Nutrients$sampledate)

Peter_Paul_Nutrients$sampledate <- 
  as.Date(Peter_Paul_Nutrients$sampledate, 
          format = "%Y-%m-%d")

###Peter_Paul_Nutrients_Gathered
colnames(Peter_Paul_Nutrients_Gathered)
class(Peter_Paul_Nutrients_Gathered$sampledate)

Peter_Paul_Nutrients_Gathered$sampledate <- 
  as.Date(Peter_Paul_Nutrients_Gathered$sampledate, 
          format = "%Y-%m-%d")

###Peter_Paul_ChemistryPhysics
colnames(Peter_Paul_ChemistryPhysics)
class(Peter_Paul_ChemistryPhysics$sampledate)

Peter_Paul_ChemistryPhysics$sampledate <- 
  as.Date(Peter_Paul_ChemistryPhysics$sampledate, 
          format = "%Y-%m-%d")

###Niwot_Litter
colnames(Niwot_Litter)
class(Niwot_Litter$collectDate)

Niwot_Litter$collectDate <- 
  as.Date(Niwot_Litter$collectDate, 
          format = "%Y-%m-%d")


```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus by phosphate, with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values.

```{r}
#need aesthetics and line of best fit
#Labels?
tp_vs_po4 <- ggplot(Peter_Paul_Nutrients, aes(y = tp_ug, x = po4)) +
  geom_point() +
  xlim(0, 50) +
  ylim(0, 150)

print(tp_vs_po4)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#Change to a factor? Or change to a date above?

#Converting month variables to factors so they can be used to break out the data in each boxplot
Peter_Paul_ChemistryPhysics$month <- as.factor(Peter_Paul_ChemistryPhysics$month)

Peter_Paul_Nutrients$month <- as.factor(Peter_Paul_Nutrients$month)

#Constructing boxplots
#Add lake as a color aesthetic
Temp_box <- ggplot(Peter_Paul_ChemistryPhysics, 
                   aes(y = temperature_C, x = month)) +
  geom_boxplot()

print(Temp_box)

tp_box <- ggplot(Peter_Paul_Nutrients, aes(y = tp_ug, x = month)) +
         geom_boxplot()
print(tp_box)

tn_box <- ggplot(Peter_Paul_Nutrients, aes(y = tn_ug, x = month)) +
         geom_boxplot()
print(tn_box)

#Create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
drymass_vs_date <- ggplot(subset(Niwot_Litter, 
                                 functionalGroup == "Needles")) +
  aes(y = dryMass, x = collectDate, color = nlcdClass) +
  geom_point()

print(drymass_vs_date)

drymass_vs_date_fac <- ggplot(subset(Niwot_Litter, 
                                     functionalGroup == "Needles")) +
  aes(y = dryMass, x = collectDate) +
  geom_point() +
  facet_wrap(vars(nlcdClass))

print(drymass_vs_date_fac)
```

Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: 