---
title: "Assignment 6: GLMs week 1 (t-test and ANOVA)"
author: "Claire Mullaney"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on t-tests and ANOVAs. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
8. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Salk_A06_GLMs_Week1.Rmd") prior to submission.

The completed exercise is due on Tuesday, February 18 at 1:00 pm.

## Set up your session

1. Check your working directory, load the `tidyverse`, `cowplot`, and `agricolae` packages, and import the NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv dataset. 

2. Change the date column to a date format. Call up `head` of this column to verify.


```{r, message = FALSE}
#1
getwd()
library(tidyverse)
library(cowplot)
library(agricolae)

Peter_Paul_Nutrients <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaul_Processed.csv")

#2
#Looking at initial column names and date format
head(Peter_Paul_Nutrients)
class(Peter_Paul_Nutrients$sampledate)

#Changing column sampledate to a date format
Peter_Paul_Nutrients$sampledate <- as.Date(Peter_Paul_Nutrients$sampledate, format = "%Y-%m-%d")

#Verifying change
class(Peter_Paul_Nutrients$sampledate)
head(Peter_Paul_Nutrients)

```


## Wrangle your data

3. Wrangle your dataset so that it contains only surface depths and only the years 1993-1996, inclusive. Set month as a factor.
```{r}
Peter_Paul_Nutrients_wr <- 
  filter(Peter_Paul_Nutrients, 
         depth == 0 & 
           year4 >= 1993 & 
           year4 <= 1996)

Peter_Paul_Nutrients_wr$month <- as.factor(Peter_Paul_Nutrients_wr$month)
class(Peter_Paul_Nutrients_wr$month)
```

## Analysis

Peter Lake was manipulated with additions of nitrogen and phosphorus over the years 1993-1996 in an effort to assess the impacts of eutrophication in lakes. You are tasked with finding out if nutrients are significantly higher in Peter Lake than Paul Lake, and if these potential differences in nutrients vary seasonally (use month as a factor to represent seasonality). Run two separate tests for TN and TP.

4. Which application of the GLM will you use (t-test, one-way ANOVA, two-way ANOVA with main effects, or two-way ANOVA with interaction effects)? Justify your choice.

> Answer: I will use a two-way ANOVA with interaction effects. We have two continuous response variables (TN and TP), each of which is being examined separately as the effect of two categorical explanatory variables, lake name and month; when two categorical variables are used as explanatory variables for a continuous response variable, a two-way ANOVA must be used in the analysis. I will consider interaction effects to account for the possibility that one of the explanatory variables depends on or affects the other. For example, if one of the lakes was in the sun and one was in the shade, levels of nitrogen or phosphorus could potentially be partially predicted by the interaction between the lake and season variables.

5. Run your test for TN. Include examination of groupings and consider interaction effects, if relevant.

>***Run the test with the interaction. Letters can be assigned to the graphs below based on the only variable that is significant, lakename. In this case, since our hypothesis was that there would be an interaction, we stopped after running run test. But model selection would also be a viable option.

>Results of lm vs. results of aov? Why aren't both lakes significant in lm?
Process for grouping when interactions are not significant but some main effects are?

6. Run your test for TP. Include examination of groupings and consider interaction effects, if relevant.

```{r}
#5 Test for tn
##As linear model
tn.anova.2way.int <- lm(data = Peter_Paul_Nutrients_wr, 
                        tn_ug ~ lakename * month)
summary(tn.anova.2way.int)

##As aov
tn.anova.2way.aov <- aov(data = Peter_Paul_Nutrients_wr, 
                         tn_ug ~ lakename * month)
summary(tn.anova.2way.aov)

##TukeyHSD post-hoc test for pairwise differences to confirm ANOVA results
TukeyHSD(tn.anova.2way.aov)

###No significant interactions; only lakename has significant differences
##No need to form groups

#6 Test for tp
##As linear model
tp.anova.2way.int <- lm(data = Peter_Paul_Nutrients_wr, 
                        tp_ug ~ lakename * month) 
summary(tp.anova.2way.int)

##As aov
tp.anova.2way.aov <- aov(data = Peter_Paul_Nutrients_wr, 
                         tp_ug ~ lakename * month) 
summary(tp.anova.2way.aov)

##TukeyHSD post-hoc test for pairwise differences to confirm ANOVA results
TukeyHSD(tp.anova.2way.aov)

##Some of the interactions, along with lakenames, are significantly different from
##each other; creating groups to more easily interpret statistical differences

##Creating an interaction as a separate variable and formatting as an aov
##to group interaction data
tp.interaction <- with(Peter_Paul_Nutrients_wr, 
                       interaction(lakename, month))
                       
tp.anova.2way.aov2 <- aov(data = Peter_Paul_Nutrients_wr, 
                          tp_ug ~ tp.interaction)
summary(tp.anova.2way.aov2)

##Groups
tp.groups <- HSD.test(tp.anova.2way.aov2, trt = "tp.interaction", 
                      group = TRUE)
tp.groups
```

7. Create two plots, with TN (plot 1) or TP (plot 2) as the response variable and month and lake as the predictor variables. Hint: you may use some of the code you used for your visualization assignment. Assign groupings with letters, as determined from your tests. Adjust your axes, aesthetics, and color palettes in accordance with best data visualization practices.

8. Combine your plots with cowplot, with a common legend at the top and the two graphs stacked vertically. Your x axes should be formatted with the same breaks, such that you can remove the title and text of the top legend and retain just the bottom legend. 

```{r}
#7
#Defining a new theme
theme_def <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")

#Setting new theme as the default theme
theme_set(theme_def)

##Plot 1 (tn)
##Groups were not created for TN, but within each month, the lake with the larger
##TN value is labelled "a" and the lake with the smaller TN value is labelled "b"
##(because the lakename significantly predicted TN concentrations)
tn.anova.plot <- ggplot(Peter_Paul_Nutrients_wr, aes(y = tn_ug, x = month, 
                                                     fill = lakename)) +
  geom_boxplot() + 
  labs(y = expression(paste("TN (", mu, "g/L)")), 
       x = "Month", fill = "Lake Name") +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62")) +
  ylim(0, 2200) +
    stat_summary(geom = "text", position=position_dodge(width=0.9), 
               fun.y = max, vjust = -1, 
               size = 3.5,
               label = c("a", "b", "a", "b", "a", 
                         "b", "a", "b", "a", "b"))

print(tn.anova.plot)

##Plot 2 (tp)
tp.anova.plot <- ggplot(Peter_Paul_Nutrients_wr, 
                        aes(y = tp_ug, x = month, 
                            fill = lakename)) +
  geom_boxplot() + 
  labs(y = expression(paste("TP (", mu, "g/L)")), 
       x = "Month", fill = "Lake Name") +
  scale_fill_manual(values = c("#66c2a5", "#fc8d62")) +
  ylim(0, 80) +
  stat_summary(geom = "text", position=position_dodge(width=0.9), 
               fun.y = max, vjust = -1, 
               size = 3.5,
               label = c("bcd", "cd", "ab", "d", "a", 
                         "d", "abc", "d", "abcd", "cd"))
  
print(tp.anova.plot)
#8
##Cowplot
tp_tn_plot <- plot_grid(tn.anova.plot + 
                          theme(legend.position = "none", 
                                axis.title.x = element_blank(),
                                axis.text.x=element_blank(),
                                axis.ticks.x=element_blank(),
                                plot.margin = margin(t = 0, r = 0, 
                                                     b = 0, l = 0,
                                                     unit = "cm")),
                        tp.anova.plot + 
                          theme(legend.position="none", 
                                plot.margin = margin(t = 0, r = 0, 
                                                     b = 0, l = 0, 
                                                     unit = "cm")),
                        nrow = 2, align = "hv")

legend <- get_legend(tn.anova.plot + 
                       guides(color = guide_legend(nrow = 1)) + 
                       theme(legend.position = "top"))


plot_grid(legend, tp_tn_plot, ncol = 1, rel_heights = c(.1, 1))

```




